---
title: "Practice Midterm: Answers"
date: "Real Thing Due: Mon Oct 18, 2021 at 11:59 PM"
output: 
  html_document: 
    fig_height: 4
    fig_width: 7
---

```{r setup, include = FALSE}
# set the echo option to FALSE to see how the document looks with the code suppressed
knitr::opts_chunk$set(echo = TRUE)
```


## Load packages

```{r load-packages, message=FALSE}
library(tidyverse)
library(nycflights13)
```

## Questions 

### Question 1

**What are the ten most common destinations for flights from NYC airports in 2013? Make a table that lists these in descending order and shows the number of fligts heading to each airport.**

```{r}
flights |>
  count(dest) |>
  arrange(desc(n)) |>
  slice(1:10)
```

### Question 2

**Which airlines have the most flights departing from NYC airports in 2013? Make a table that lists these in descending order of frequency and shows the number of flights for each airline. In your narrative mention the names of the airlines as well. *Hint:* You can use the `airlines` dataset to look up the airline name based on `carrier` code.***

Prof Note: We have not discussed `inner_join` yet...this will not show up on your real midterm.

```{r}
flights |>
  count(carrier) |>
  arrange(desc(n)) |>
  inner_join(airlines, by = "carrier")
```

The carrier with the highest number of flights departing from NYC airports in 2013 is United Airlines, followed by JetBlue Airways and ExpressJet Airlines.

### Question 3

**Consider only flights that have non-missing arrival delay information. Your answer should include the name of the carrier in addition to the carrier code and the values asked. (1a) Which carrier had the highest mean arrival delay?(1b) Which carrier had the lowest mean arrival delay? Make sure that your answer includes the name of the carrier and the calculated mean delay.**

```{r}
flights |>
  filter(!is.na(arr_delay)) |>
  group_by(carrier) |>
  summarise(mean_arr_delay = mean(arr_delay)) |>
  arrange(desc(mean_arr_delay)) |>
  inner_join(airlines, by = "carrier") |>
  slice(c(1, n()))
```

Highest mean arrival delay was on Frontier Airlines with 21.9 minutes, and lowest mean was on Alaska Airlines with -9.93 minutes (which means 9.93 minutes early on average).

### Question 4

**What was the mean temperature at the origin airport on the day with the highest departure delay? Your answer should include the name of origin airport, the date with the highest departure delay, and the mean temperature on that day.**

```{r}
flights |>
  arrange(desc(dep_delay)) |>
  slice(1) |>
  select(dep_delay, month, day, origin)
```

The highest departure delay of 1301 minutes (approx 21.7 hours) was on a flight departing from JFK on Jan 9. 

```{r}
weather |>
  filter(month == 1, day == 9, origin == "JFK") |>
  summarise(mean_temp = mean(temp))
```

The average temperature on this day was 40.835 F.

### Question 5

**Consider breaking the day into four time intervals: 12:01am-6am, 6:01am-12pm, 12:01pm-6pm, 6:01pm-12am. (a) Calculate the proportion of flights that are delayed at departure at each of these time intervals. (b) Comment on how the likelihood of being delayed change throughout the day?**

```{r}
# Create time of day variable
flights <- flights |>
  mutate(time_of_day = case_when(
    sched_dep_time >= 001  & sched_dep_time <= 600  ~ "12:01am-6am",
    sched_dep_time >= 601  & sched_dep_time <= 1200 ~ "6:01am-12pm",
    sched_dep_time >= 1201 & sched_dep_time <= 1800 ~ "12:01pm-6pm",
    sched_dep_time >= 1801                          ~ "6:01pm-12am"
  ))

# Find proportion of delayed flights for each time of day
flights |>
  filter(!is.na(dep_delay)) |>
  mutate(dep_delayed = case_when(dep_delay > 0 ~ "delayed", 
                                 dep_delay <= 0 ~ "ontime")) |>
  count(time_of_day, dep_delayed) |>
  group_by(time_of_day) |>
  mutate(prop_delayed = n / sum(n)) |>
  filter(dep_delayed == "delayed") |>
  arrange(prop_delayed)
```

(a) Approx 21% of flights are delayed between midnight and 6am, 26% of flights are delayed between 6am and noon, 46% of flights are delayed between noon and 6pm, and 52% of flights are delayed between 6pm and midnight.
(b) As the day progresses, the likelihood of being delayed at departure increases.

### Question 6

**Find the flight with the longest air time. (a) How long is this flight? (b) What city did it fly to? (c) How many seats does the plane that flew this flight have?**

```{r}
flights |>
  arrange(desc(air_time)) |>
  slice(1) |>
  select(air_time, dest, tailnum) |>
  inner_join(planes, by = "tailnum") |>
  select(air_time, dest, tailnum, seats)
```

(a) The flight time is 695 minutes (11.58 hours).
(b) Flight was to Honolulu. 
(c) Plane that flew this flight has 292 seats.

### Question 7

**The `airports` data frame contains information on a large number of primarily American airports. These data include location information for these airports in the form of latitude and longitude coordinates. In this question we limit our focus to the [Contiguous United States](https://en.wikipedia.org/wiki/Contiguous_United_States). Visualize and describe the distribution of the longitudes of airports in the Contiguous United States. What does this tell you about the geographical distribution of these airports? *Hint:* You will first need to limit your analysis to the Contiguous United States. [This Wikipedia article](https://en.wikipedia.org/wiki/List_of_extreme_points_of_the_United_States) can help, but you're welcomed to use other resources as well. Make sure to cite whatever resource you use.**

Based on information from [this Wikipedia article](https://en.wikipedia.org/wiki/List_of_extreme_points_of_the_United_States) as well as [this conversion site](https://www.vercalendario.info/en/how/convert-latitude-longitude-degrees-decimals.html), we use the following points as the boundaries of the 48 Contiguous United States:

- Northernmost point in the 48 contiguous states: Northwest Angle Inlet in Lake of the Woods, Minnesota 49°23′04.1″N 95°9′12.2″W 
    - lat: 49.38447222222222
    - lon: -95.1533888888889
- Southernmost point in the 48 contiguous states continuously above water: Ballast Key, Florida (24°31′15″N 81°57′49″W)
    - lat: 24.520833333333332
    - lon: -81.96361111111112
- Westernmost point in the 48 contiguous states continuously above water: Bodelteh Islands, offshore from Cape Alava, Washington 48°10′42.7″N 124°46′18.1″W
    - lat: 48.178527777777774 
    - lon: -124.77169444444445 
- Easternmost point in the 50 states: Sail Rock, Maine 44°48′45.2″N 66°56′49.3″W
    - lat: 44.812555555555555
    - lon: -66.94702777777778

```{r}
airports |>
  filter(
    lat > 24.520833333333332 & lat < 49.38447222222222,
    lon > -124.77169444444445 & lon < -66.94702777777778
  ) |>
  ggplot(aes(x = lon)) +
    geom_histogram(binwidth = 5)
```

The distribution of longitudes is bimodal. One of the modes is at latitude -120, correspomding to the West Coast and the other is around latitude -85 corresponding to the midtwest.

### Question 8

**Recreate the plot included below using the `flights` data. Once you have created the visualization, in no more than one paragraph, describe what you think the point of this visualization might be. *Hint:* The visualization uses the variable `arrival`, which is not included in the `flights` data frame. You will have to create `arrival` yourself, it is a categorical variable that is equal to `"ontime"` when `arr_delay <= 0` and `"delayed"` when `arr_delay > 0`.**

```{r}
flights |> 
  filter(month == 12, dest %in% c("PHL", "RDU")) |> 
  mutate(arrival = case_when(arr_delay > 0 ~ "delayed", 
                             arr_delay <= 0 ~ "ontime")) |>
  filter(!is.na(arrival)) |>
  ggplot(aes(x = arrival, y = dep_delay, color = dest)) + 
    geom_boxplot() + 
    facet_grid(dest ~ origin) + 
    labs(title = "On time performance of NYC flights",
         subtitle = "December 2013",
         x = "Arrival",
         y = "Departure delay",
         color = "Destination")
```
