---
title: "13-cs01-analysis"
author: "Professor Shannon Ellis"
date: "2023-11-14"

format:
  html: 
    output-file: 13-cs01-analysis.html
    embed-resources: true
  revealjs:
    output-file: 13-cs01-analysis-slides.html
    slide-number: true
    chalkboard: false 
    preview-links: auto
    logo: images/cogs137-logo-hex.png
    css: slides.css
    footer: <https://cogs137.github.io/website/>
    scrollable: true
    embed-resources: true
    execute:
      echo: true
      eval: true
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(emo)

knitr::opts_chunk$set(dpi=300, echo=TRUE, warning=FALSE, message=FALSE) 
```

# CS01: Biomarkers of Recent Use (Analysis) {background-color="#92A86A"}

## Q&A {.smaller}

> Q: \
> A: 


## Course Announcements

Due Dates:

-   `r emo::ji("science")` No Lab this week (holiday) - all students will receive full credit for Lab05 (MLR)
-   `r emo::ji("question")` **Mid-course survey** "due" (for EC) Friday
-   `r emo::ji("computer")` HW03 (MLR) due Mon 11/20


## Agenda

-   

# Question {background-color="#92A86A"}

Which compound, in which matrix, and at what cutoff is the best biomarker of recent use?

# Data & Files {background-color="#92A86A"}

## Packages

Two additional packages required for these notes:

```{r}
library(tidyverse)
library(purrr)
library(rstatix)
```

## The Data

Reading in the data from the end of data wrangling notes: 

```{r}
load("data/compounds.RData")
load("data/timepoints.RData")
load("data/data_clean.RData")
```

```{r, echo=FALSE,eval=FALSE}
load("content/lectures/data/compounds.RData")
load("content/lectures/data/timepoints.RData")
load("content/lectures/data/data_clean.RData")

```

# Analysis {background-color="#92A86A"}



## Sensitivity & Specificity

```{r}

```


## Recap {.smaller background-color="#92A86A"}

-   Can you explain/describe the plots generated in the context of these data?
-   Can you generate EDA plots of your own for these data
-   Can you undersetand/work through the more complicated code provided (even if you couldn't have come up with it on your own)
