---
title: "16-cs02-eda"
author: "Professor Shannon Ellis"
date: "2023-03-07"

format:
  html: 
    output-file: 16-cs02-eda.html
    embed-resources: true
  revealjs:
    output-file: 16-cs02-eda-slides.html
    slide-number: true
    chalkboard: false 
    preview-links: auto
    logo: images/cogs137-logo-hex.png
    css: slides.css
    footer: <https://cogs137.github.io/website/>
    scrollable: true
    embed-resources: true
    execute:
      echo: true
      eval: true
---

# CS02: Vaping Behaviors in American Youth (EDA) {background-color="#92A86A"}



## Course Announcements {.smaller}

**Due Dates**:

- Lecture Participation survey "due" after class
- Lab08 Due Friday (CS02 EDA)
- CS02 due Monday of Finals week (3/20)
- Final Project due Th of Finals week (3/23)

. . .

**Notes**:

- cs01 group work form is still open - please complete (if you haven't)
- cs02 and final project groups/repos assigned
- don't wait until finals week to do cs02 and final project



```{r packages, echo = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")

knitr::opts_chunk$set(fig.height = 3.5, dpi = 300, echo=TRUE, warning=FALSE, message=FALSE) 
```

## Agenda


# Questions {background-color="#92A86A"}

1. How has tobacco and e-cigarette/vaping use by American youths changed since 2015?
2. How does e-cigarette use compare between males and females?
3. What vaping brands and flavors appear to be used the most frequently?
4. Is there a relationship between e-cigarette/vaping use and other tobacco use?


## Data

...will only work if you finished last set of notes.

```{r, eval=FALSE}
load("data/wrangled/wrangled_data_vaping.rda")
```

## Question

When is pivoting your data from wide to long (or long to wide) helpful?

. . . 

```{r, eval=FALSE}
data |>
   pivot_longer(cols = columns_to_pivot , names_to = "new_col_for_labels" , values_to = "new_col_for_values")
```


## EDA

What do you want to know? 

\[We're going to fill these in during class\]

. . .

How would you go about doing that?



## Suggested Reading

-   [Case Study from OCS](https://www.opencasestudies.org/ocs-bp-vaping-case-study/)
