---
title: "Lab 02: Wrangling"
author: ""
format:
  revealjs:
    slide-number: true
    chalkboard: false 
    preview-links: auto
    logo: images/cogs137-logo-hex.png
    css: slides.css
    footer: <https://cogs137.github.io/website/>
    execute:
      echo: true
---

# Lab 02: Wrangling {background-color="#0E3A53"}

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Reminders

- Start with `library(tidyverse)` (includes `tidyr`, `readr`, `dplyr`, etc.)
- Clone using 'SSH' link from GitHub
- Knit to .html & push both .Rmd and .html to GitHub

## Agenda

1.  **Lab 02 intro and demos:** Introduce the lab, and work through the first question as a class.
2.  **On your own:** Work on the rest of the lab "on your own", but feel free to check in with classmates as much as you like.

## `dplyr`: Review {.scrollable .smaller}

**dplyr** provides a "Grammar of Data Manipulation" and is based on the concepts of functions as verbs that manipulate data frames.

-   `filter`: pick rows matching criteria
-   `slice`: pick rows using index(es)
-   `select`: pick columns by name
-   `pull`: grab a column as a vector
-   `rename`: rename specific columns
-   `arrange`: reorder rows
-   `mutate`: add new variables
-   `transmute`: create new data frame with variables
-   `distinct`: filter for unique rows
-   `sample_n` / `sample_frac`: randomly sample rows
-   `summarize`: reduce variables to values
-   ... (many more)

## The Data {.smaller}

```{r}
storms |>
  slice(1:20)
```

## The Data: Documentation

From the console...

```{r}
?storms
```
