---
title: "Lab 03: Exploring & Visualizing Data"
author: ""
format:
  revealjs:
    slide-number: true
    chalkboard: false 
    preview-links: auto
    logo: images/cogs137-logo-hex.png
    css: slides.css
    footer: <https://cogs137.github.io/website/>
    execute:
      echo: true
---

# Lab 03: Exploring & Visualizing Data {background-color="#0E3A53"}


```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Agenda

1. **Lab introduction:** Review FiveThirtyEight article on college majors.
2. **Tips:** 
- Review a few common RMarkdown mistakes.
- Review continuous vs discrete variable types


## Lab Intro

- Lab instructions posted on the course website.

- [The Economic Guide To Picking A College Major](https://fivethirtyeight.com/features/the-economic-guide-to-picking-a-college-major/) by Ben Casselman
    
## Reminders

-   Start with `library(tidyverse)` (includes `tidyr`, `readr`, `dplyr`, etc.)
-   Clone using 'SSH' link from GitHub
-   Knit to .html & push both .Rmd and .html to GitHub

## Tips {.scrollable}

- Be ready to troubleshoot your document, since it will likely fail to knit on multiple occasions throughout the process. Read the error message carefully and take note of which line is preventing a successful knit.

- Make sure to keep track of your various chunks and to keep text and code in the right place.

- Remember that your R Markdown file is not aware of your project's global environment and can only make use of variables, functions, etc. that you have loaded or defined in the document.

- Remind yourself how the pipe operator (`|>`) works.

- If you're unsure how a function works or what its arguments are, type `?` in front of it and hit enter (`?read_csv` for instance). The "Help" tab will open and provide a summary of the function as well as some examples.

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

# Demo {background-color="#0E3A53"}