---
title: "Lab 01: Tooling"
author: ""
date: "2021-10-01"
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

class: center, middle

# Welcome to lab

---

## Agenda

1. **Lab strucuture:** Lab structure overview.
2. **Lab 01 intro and demos:** Introduce the lab, and work through the first section as a class.
3. **On your own:** Work on the rest of the lab "on your own", but feel free to check in with classmates as much as you like.

---

class: center, middle

# Lab strucuture

---

## Lab strucuture

- 5-10 minute introduction (a bit longer today)

- Use the remaining time to work through the lab exercises and fill out your lab report
    + Submit: on your own
    + Working: always allowed to work together

- Lab instructions posted on the course website under `</>`
    + Let's go find today's lab!
    
---

## Tips

- You do not have to finish the lab in class; you have until midnight to submit. But, you might choose to get through portions that you think will be challenging (which initially might be the coding component) in class when staff can help you on the spot, and leave the narrative writing until later.

- Do not pressure each other to finish early; use the time wisely to really learn the material and produce a quality lab

- When working with others, DO not split up lab among classmates, work on it together in its entirety.

---

class: center, middle

# Lab 01 intro and demos

---

## Goals

- Introduce you to R and RStudio:
    + R is the name of the programming language itself 
    + RStudio is a convenient interface

--

- Introduce you to Git and GitHub: collaboration and version control system that we will be using throughout the course
    + Git is a version control system -- like "Track Changes" features from Microsoft Word on steroids
    + GitHub is the home for your Git-based projects on the internet -- like DropBox but much, much better

---

## Getting started

First, put away computers, and watch me do it:

- Click on the assignment link on Canvas for today's lab to create your GitHub repository (which we’ll refer to as "repo" going forward) for the lab. This repo contains a template you can build on to complete your lab.

- On GitHub, accept the assignment. Click on the clipboard icon to copy the repo URL.

- Go to datahub. and open RStudio. Go to **File > New Project**... and select to create a **New Project from Version Control**. On the following menu, select **Git**. 

- Copy and paste the URL of your assignment repo into the "Repository URL" dialog box.

- Hit **Create Project**.

Now it's your turn! Place a green sticky on your laptop when you're done with this part (you can continue if you like). Place a pink sticky if you have questions.

---

## Housekeeping

First, put away computers, and watch me do it:

- Go to the Terminal pane

- Type the following two lines of code, replacing the information in the quotation marks with your info:

```
git config --global user.email "your email"
git config --global user.name "your name"
```

- To confirm that the changes have been implemented, run the following

```
git config --global user.email
git config --global user.name
```

Now it's your turn! Place a green sticky on your laptop when you're done with this part (you can continue if you like). Place a red sticky if you have questions.

---

class: center, middle

# On your own!

<iframe src="https://giphy.com/embed/13awhIC2CcDaHC" width="480" height="336" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/fall-kid-childhood-13awhIC2CcDaHC">via GIPHY</a></p>
