---
title: "Lab 05: Exploring CS1 Youth Disconnection"
author: ""
date: "2021-10-22"
output: 
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.35, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Lab 05: Exploring Case Study 1

- Using same data we've been discussing in class
- Provided in an already-wrangled format
- For your hw03 and case study report, you will have to do the wrangling
- but in this lab, the goal is to get started working with the data

---

## Reminder that this is based off of...

"Beauty in the classroom: instructors’ pulchritude and putative pedagogical productivity" 

Wright, Carrie and Ontiveros, Michael and Jager, Leah and Taub, Margaret and Hicks, Stephanie C. (2020). https://github.com/opencasestudies/ocs-youth-disconnection-case-study. Disparities in Youth Disconnection.

---

## Some new challenges in this lab
- Little guidance for Part 1 - it's up to you to determine what visualizations help you explore and learn about the data; feel free to discuss with classmates!
- There are some unwieldy functions you'll be looking at and explaining in Part 2 - it's the first time you've done something like this. Take the code line-by-line if you're struggling to digest it all at once
- Brainstorming in Part 3 is meant to be *very* open. There are no "wrong" ideas or "bad" suggestions. If you've got a thought, include it there. For your final report and in your discussions with your group, you can narrow your ideas down. Better to start with too many ideas than not enough!




