---
title: "Data Import & Wrangling I"
author: "Prof Ellis"
date: "2021-10-08"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magick)
library(knitr)
library(emo)
library(mosaicData)
library(openintro)
library(DT)
```

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}
```

<!-- Add in information about coding variables -->

## Announcements

- Class recordings linked on course page -- I'll email the link as well
- Look out for an email about your team tonight, start thinking about a team name asap
- 

---

class: center, middle

# Tidy data

---

## Tidy data

>Happy families are all alike; every unhappy family is unhappy in its own way. 
>
>Leo Tolstoy

--

.pull-left[
**Characteristics of tidy data:**

1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table.
]
--
.pull-right[
**Characteristics of untidy data:**

!@#$%^&*()
]

---

## Summary tables 

.question[
Is each of the following a dataset or a summary table?
]

.small[
.pull-left[
```{r echo=FALSE}
starwars %>%
  select(name, height, mass) %>%
  datatable(options = list(dom = "pt", pageLength = 8))
```
]
.pull-right[
```{r echo=FALSE}
starwars %>%
  count(homeworld, gender) %>%
  spread(gender, n) %>%
  datatable(options = list(dom = "pt", pageLength = 8))
```
]
]

---



## What about other arguments?

To send results to a function argument other than first one or to use the previous result for multiple arguments, use `.`:

```{r fig.height=1.8, fig.width = 3.75}
starwars %>%
  filter(species == "Human") %>%
  lm(mass ~ height, data = .)
```

---


