---
title: "CS1: Youth Connection (EDA)"
author: "Prof Ellis"
date: "2021-10-27"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    anchor_sections: FALSE
    nature:
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false
---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)

# For nonsese...
library(emo)
library(DT)
```

```{r child = "setup.Rmd"}
```

## Q&A

.small[
Q: When running the command, install.packages("pdftools"), I am getting an error.  
A: This should now be fixed! `pdftools`, `magick`, and `tesseract` have now been installed for you on datahub. You can just load them in with `library()` and get started. Apologies!
]

---
## Course Announcements

**Case Study Details**:
- This Friday (10/29): 
  - lab05 - exploring CS1
  - [Group Submission form](https://docs.google.com/forms/d/e/1FAIpQLSf-ASyUBp10SnyOp6bB7GuKwRLSaeg86KczpT1m7ExINbvbew/viewform?usp=sf_link) (link also on canvas)
- This Coming Monday (11/1): hw03 - CS1 Rough Draft Due
- The following Monday (11/8): CS1 Full Report Due

[**Mid-course Survey**](https://docs.google.com/forms/d/e/1FAIpQLScyO_RW1yfvVtDX160kbzh2dwDaD_4EI4HaQWaYN0uueWve2A/viewform?usp=sf_link) (*optional*; due Mon 11/1 for extra credit)

---

## The Data

This is where we left off in the last set of notes...

```{r, message = FALSE}
major_groups_long <- read_csv("data/wrangled_major_groups_data.csv")
asian_subgroups <- read_csv("data/wrangled_asian_subgroups_data.csv")
latinx_subgroups <- read_csv("data/wrangled_latinx_subgroups_data.csv")
```

---
## The Data 
.small[

```{r, echo=FALSE}
DT::datatable(major_groups_long)
```
]

---

## EDA: Disonnection over time

.panelset[

.panel[.panel-name[Code]
```{r eda-overall, fig.show = "hide"}
major_groups_long |>
  filter(Gender == "All", Race_Ethnicity == "All_races") |>
  ggplot(aes(x = Year, y = Percent)) +
    geom_line(size = 0.5) +
    geom_point(size = 3) 
```
]

.panel[.panel-name[Plot]
```{r ref.label="eda-overall", echo = FALSE, out.width="100%"}
```
]

.panel[.panel-name[Notes]
Notes to self:
- increases initially in 2010; decreases since then
- x-axis handled as an integer 
- no informative title
- stylistic improvements can be made
]
]

---

## EDA: Disconnection by Gender

.panelset[

.panel[.panel-name[Code]
```{r eda-gender, fig.show = "hide"}
major_groups_long |>
  filter(Gender != "All", Race_Ethnicity == "All_races") |>
  ggplot(aes(x = Year, y = Percent, color = Gender)) +
    geom_line(size = 0.5) +
    geom_point(size = 3) 
```
]

.panel[.panel-name[Plot]
```{r ref.label="eda-gender", echo = FALSE, out.width="100%"}
```
]

.panel[.panel-name[Notes]
Notes to self:
- similar overall patterns; most years Male is higher than Female
- colors likely not final colors we would want to use
- directly label lines?
- title/style issues remain
]
]

---

## EDA: Disconnection by Race



.panelset[

.panel[.panel-name[Code]
```{r eda-race, fig.show = "hide"}
major_groups_long |>
  filter(Gender == "All", Group != "United States") |>
  ggplot(aes(x = Year, y = Percent, color = Race_Ethnicity)) +
    geom_line(size = 0.5) +
    geom_point(size = 3) 
```
]

.panel[.panel-name[Plot]
```{r ref.label="eda-race", echo = FALSE, out.width="100%"}
```
]

.panel[.panel-name[Notes]
Notes to self:
- definite differences by race/ethnicity
- colors likely not final colors we would want to use
- directly label lines
- title/style issues remain
]
]


---

## Other Data

We haven't even begun to explore the subgroups data (`latinx_subgroups`, `asian_subgroups`) nor the intersection of gender and ethnicity yet...might be something worth exploring during lab!
