---
title: "Effective Communication & Wrap Up"
author: "Prof Ellis"
date: "2021-12-3"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r child = "setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# For nonsese...
library(emo)
library(DT)
library(tidyverse)

knitr::opts_chunk$set(warning=FALSE, message=FALSE) 
```

**Q&A**

.small[
Q: Within R/Python, when should one use SQL over dplyr/Pandas?  
A: SQL when you need to access information stored in a relational database. Note that there is an R package (`RSQLite`) that helps you do that from R.


Q: In Pandas, is there a similar operator like the Pipe operator in R to chain operations?  
A: No. You can use `.function().funtion()` to chain operators together, but there is nothing that allows for the same ordering/piping in pandas. There is a package that attempts to do so, but it's not fully developed yet. Another note is that there is a package `reticulate` that allows you to execute Python code from R, so you could always combine the two.

Q: Is there a way to save the "website" repository?  
A: You could fork it.

Q: The plot looks a lot like those would be generated in tableau especially the us map one  
A: Yup - a lot of the more "advanced" visualizations (especially r2d3) often look like what would be generated from software like Tableau!
]

---

## Course Announcements

- **cs03** due tonight (12/3) 
- **final project due Wednesday** (12/8; 11:59 PM) 
- report + group work form for cs03 and final project (link on Canvas home page)

--

- [Lecture Slide .Rmds](https://github.com/COGS137/website/tree/master/static/lecslides)
- lab08 graded; lab09 and hw05 "graded"
- accessing cs03 and final project repos: https://github.com/COGS137

--

- Please fill out your [CAPEs](https://cape.ucsd.edu/)! (currently: 40%)
- [Post-course survey](https://docs.google.com/forms/d/e/1FAIpQLScBKhtzUSAJRraefPIr1K6pZEnyVdROWle-Ag_sgyC8cQXwlw/viewform?usp=sf_link)
- [OCS Survey](https://www.opencasestudies.org/) - one submission per case study

---

class: center, middle

## CS02 Discussion

What was your experience like completing CS02 (RTC and violent crime)?

---

## Effective Written Communication

What are some tips for effective written communication?


- Hold the reader's attention with what needs to be said; do so effectively
- consider your audience
- Know how much to write? Concise; showing what needs to be shown 
- tell your audience why they should care; why it matters
- go from general to more specific
- defend/explain your choices
- proofread 

---

## vs. Oral Communication

- repetition is ok: tell what you're going to tell them, tell them, tell them what you told them
- out loud you have more room for error b/c you're there to contextualize 
- viz: more on reports; less busy viz on oral presentations
- less text on slides 

---

### Storytelling

What makes a good story?

- introduction (beginning), conflict/action (middle), climax/resolution (end)
- as interesting as possible
- clear 

---

### Background

What should be included in a good background section?

- definitions of unknown/important terms - intro the main topics
- relevant literature - talk about/summarize important findings; tie back to current work

---

#### Background: Practice

Topic: What is the relationship between healthcare expenditures and health outcomes in the United States?

- Health care in the United States (insurance, patients, etc.)
- What are healthcare expenditures/how being defined/what included
- health outcomes - explain; define; 
- Prior research/work on this topic - summarize; contextualize; why does their work matter here?

---

#### Citations: When Needed?

.small[
"We will be doing our analysis using two different data sets created by two different groups: Donohue and Mustard + Lott, or simply Lott"
]
--

.small[
"What turned from the idea of carrying firearms to protect oneself from enemies such as the British monarchy and the unknown frontier of North America has now become a nationwide issue" 
]
--

.small[
"Right to Carry Laws refer to laws that specify how citizens are allowed to carry concealed handguns when they’re away from home without a permit"
]
--

.small[
"In this case study, we are examining the relationship between unemployment rate, poverty rate, police staffing, and violent crime rate."
]
--

.small[
"In the United States, the second amendment permits the right to bear arms, and this law has not been changed since it’s creation in 1791"
]
--

.small[
"The Right to Carry Laws (RTC) is defined as “a law that specifies if and how citizens are allowed to have a firearm on their person or nearby in public.”"
]

--

You do NOT get docked points for citing others' work. You can be at risk of AI Violation if you don't. **When in doubt, give credit.**

---
#### Aside: Footnotes

Possible to specify footnotes in Rmd documents:

Here is some body text.[^1]

[^1]: This footnote will appear at the bottom of the page.

--

**Superscript**

Superscript also possible: <sup>1</sup> (`^1^`)

---

### Data Section

What should be included in a data section?

- Description of dataset and variables
- Wrangling
- Show a portion of the data
- EDA (could also be in analysis section)

---

class: center, middle

## CS02: Feedback

---

## CS02: Feedback (Narrative)

Report should tell a narrative. 
- You can't refer to something that hasn't been introduced
- Should make sense when read from top to bottom

--

- Code should be introduced
  - General guidance in text
  - specific code explanations as code comments
    
---

## CS02: Feedback (Background)

Background section should:
  - Introduce the topic(s)
  - Include relevant context
  - Include references/citations as needed

---

## CS02: Feedback (Questions)

Questions posed should all be included in questions section
- for case studies, the question(s) posed in class
- and any questions you answer in extension

---

## CS02: Feedback (Analysis)

- EDA should *fully* explore your datasets; not just what was presented in class

--

- Text included should:
  - not *only* explain what you did
  - should explain what you found
  - should explain what your findings mean *in context of questions being answered*
  
--

- Analyses should answer your questions - double check this

--

- Visualizations should:
  - not just explain what you plotted
  - accompanying text should explain what reader shoudl learn
  - should contextualize what that means for question(s) being answered

---

### CS03 + Final Project

- Narrative/Story
- Pertinent background
- Stated questions
- Clear explanations
- Clear (& commented, where necessary) code
- Accurate and Interpreted Viz + Analysis

--

.question[Questions about CS03 + Final Project Tasks or Expectations?]

---

## COGS 137: Where We've Been

- R, RMarkdown & RStudio
- Data Wrangling w/ the `tidyverse`
- Dataviz w/ `ggplot2`
- CS01: Linear Models + Youth Connection
- CS02: MLR, Multicollinearity & Right to Carry
- CS03: Logistic Reression & Youth Vaping
- Next Steps in R: `Shiny`, `bookdown`, `blogdown`, `plotly`/`gganimate`

---

## COGS 137: A Brand New Course

**Lots of thanks!**

- course staff! (Sean, Shubham, & Colin - feedback, grading, answer keys, labs, office hours, etc.)

--

- all of you (hard work, patience as deadlines moved, flexibility, interest, feedback)

--

- [Mine Çetinkaya-Rundel](https://mine-cr.com/), [Open Case Studies](https://www.opencasestudies.org/) Team, RStudio & tidyverse team 
- Sean Kross & Prof Drew Walker

---

class: center, middle

## Good Luck on Finals, Get Sleep, Be Safe, Drink Water, Take Care of Yourselves, & Have a Wonderful Break!