---
author: "Prof Ellis"
date: "2021-11-10"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r child = "setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# For nonsese...
library(emo)
library(DT)

knitr::opts_chunk$set(warning=FALSE, message=FALSE) 
```

## Course Announcements

- **lab07**: "due Friday" but accepted until Sunday night (multiple linear regression)
- **hw04** due this Monday 11/15 (individually; wrangling of subset of data)
- **cs02** due next Monday 11/22 
- heads up: No class Wed of Week 9 (day before Thanksgiving)

---

## Question

.question[What is the effect of multicollinearity on coefficient estimates from linear regression models when analyzing right to carry laws and violence rates?]

---

# Packages 

```{r packages-load}
library(tidyverse)
library(skimr) # will need to install first
library(ggrepel)
```

---

## Data

```{r}
load("data/wrangled/wrangled_data_rtc.rda")
```

---

### LOTT data

.panelset[
.panel[.panel-name[glimpse]
.small[
```{r}
glimpse(LOTT_DF)
```
]]
.panel[.panel-name[skim]
.small[
```{r}
skimr::skim(LOTT_DF)
```
]]]


---

### DONOHUE data

.panelset[
.panel[.panel-name[glimpse]
.small[
```{r}
glimpse(DONOHUE_DF)
```
]]
.panel[.panel-name[skim]
.small[
```{r}
skimr::skim(DONOHUE_DF)
```
]]]

---
## Population over time
.panelset[
.panel[.panel-name[Plot]

```{r ref.label="population", echo = FALSE, out.width="100%"}
```
]
.panel[.panel-name[Code]
```{r population}
DONOHUE_DF |>
  group_by(YEAR) |>
  summarise(Population = sum(Population)) |> 
  ggplot(aes(x = YEAR, y = Population)) +
  geom_line() +
  scale_x_continuous(
    breaks = seq(1980, 2010, by = 1),
    limits = c(1980, 2010),
    labels = c(seq(1980, 2010, by = 1))
  ) +
  labs(
    title = "Population has steadily increased",
    x = "Year",
    y = "Population"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90),
        plot.title.position = "plot")
```
]]
---

## Crime over time

.panelset[
.panel[.panel-name[Plot]

```{r ref.label="donohue-overall", echo = FALSE, out.width="100%"}
```
]
.panel[.panel-name[Code]
.very-small[
```{r donohue-overall, fig.show = "hide"}
DONOHUE_DF |>
  group_by(YEAR) |>
  summarize(
    Viol_crime_count = sum(Viol_crime_count),
    Population = sum(Population),
    .groups = "drop"
  ) |>
  mutate(Viol_crime_rate_100k_log = log((Viol_crime_count * 100000) / Population)) |>
  ggplot(aes(x = YEAR, y = Viol_crime_rate_100k_log)) +
  geom_line() +
  scale_x_continuous(
    breaks = seq(1980, 2010, by = 1),
    limits = c(1980, 2010),
    labels = c(seq(1980, 2010, by = 1))
  ) +
  scale_y_continuous(
    breaks = seq(5.75, 6.75, by = 0.25),
    limits = c(5.75, 6.75)
  ) +
  labs(
    title = "Crime rates fluctuate over time",
    x = "Year",
    y = "ln(violent crimes per 100,000 people)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90),
        plot.title.position = "plot")
```
]]
]

---

## Crime over time by state

.panelset[
.panel[.panel-name[Plot]

```{r ref.label="donohue-state", echo = FALSE, out.width="100%"}
```
]
.panel[.panel-name[Code]
.very-small[
```{r donohue-state, fig.show = "hide"}
DONOHUE_DF |>
  mutate(Viol_crime_rate_100k_log = log((Viol_crime_count * 100000) / Population)) |>
  ggplot(aes(x = YEAR, y = Viol_crime_rate_100k_log, color = STATE)) +
  geom_point(size = 0.5) +
  geom_line(aes(group = STATE),
    size = 0.5,
    show.legend = FALSE
  ) +
  geom_text_repel(data = DONOHUE_DF |>
      mutate(Viol_crime_rate_100k_log = log((Viol_crime_count * 100000) / Population)) |>
      filter(YEAR == last(YEAR)),
      aes(label = STATE,x = YEAR, y = Viol_crime_rate_100k_log),
      size = 3, alpha = 1, nudge_x = 1, direction = "y",
      hjust = 1, vjust = 1, segment.size = 0.25, segment.alpha = 0.25,
      force = 1, max.iter = 9999) +
  guides(color = "none") +
  scale_x_continuous(
    breaks = seq(1980, 2015, by = 1),
    limits = c(1980, 2015),
    labels = c(seq(1980, 2010, by = 1), rep("", 5))
  ) +
  scale_y_continuous(
    breaks = seq(3.5, 8.5, by = 0.5),
    limits = c(3.5, 8.5)
  ) +
  labs(
    title = "States have different levels of crime",
    x = "Year",
    y = "ln(violent crimes per 100,000 people)"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90), plot.title.position = "plot")
```
]]
]

---

## Police Presence over time

.panelset[
.panel[.panel-name[Plot]

```{r ref.label="police", echo = FALSE, out.width="100%"}
```
]
.panel[.panel-name[Code]
```{r, police}
DONOHUE_DF |>
  group_by(YEAR) |>
  summarise(Police = sum(police_per_100k_lag)) |> 
  ggplot(aes(x = YEAR, y = Police)) +
  geom_line() +
  scale_x_continuous(
    breaks = seq(1980, 2010, by = 1),
    limits = c(1980, 2010),
    labels = c(seq(1980, 2010, by = 1))
  ) +
  labs(
    title = "Police Presence has increased over time with fluctuations",
    x = "Year",
    y = "Police Presence per 100K people"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90),
        plot.title.position = "plot")
```
]
]
