---
title: "CS03 (Data)"
author: "Prof Ellis"
date: "2021-11-18"
output:
  xaringan::moon_reader:
    css: "slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---


```{r child = "setup.Rmd"}
```

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
# For nonsese...
library(emo)
library(DT)
library(tidyverse)

knitr::opts_chunk$set(warning=FALSE, message=FALSE) 
```


## Q&A



---

## Course Announcements  

- **cs02** due  Monday
- **cs03** due 
--

- For cs03 and final project, there will be a form for each individual to fill out about how group work went; this can be used to adjust individual's scores. 
---

## Questions

1. How has tobacco and e-cigarette/vaping use by American youths changed since 2015?
2. How does e-cigarette use compare between males and females?
3. What vaping brands and flavors appear to be used the most frequently?
4. Is there a relationship between e-cigarette/vaping use and other tobacco use?

---

## Load the Data

...will only work if you finished last set of notes.

```{r, eval=FALSE}
load("data/wrangled/wrangled_data_vaping.rda")
```



