---
title: "Lab 02: Data Wrangling"
author: ""
date: "2021-10-08"
output:
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.65, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
library(DT)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```


## Agenda

1. **Lab 02 intro and demos:** Introduce the lab, and work through the first question as a class.
2. **On your own:** Work on the rest of the lab "on your own", but feel free to check in with classmates as much as you like.

---

## `dplyr`: Review

**dplyr** provides a "Grammar of Data Manipulation" and is based on the concepts of functions as verbs that manipulate data frames.

.midi[
* `filter`: pick rows matching criteria
* `slice`: pick rows using index(es)
* `select`: pick columns by name
* `pull`: grab a column as a vector
* `rename`: rename specific columns
* `arrange`: reorder rows
* `mutate`: add new variables
* `transmute`: create new data frame with variables
* `distinct`: filter for unique rows
* `sample_n` / `sample_frac`: randomly sample rows
* `summarise`: reduce variables to values
* ... (many more)
]

---

## The Data

.small[
```{r}
storms |>
  slice(1:20)
```
]

---

## The Data: Documentation

From the console...

```{r}
?storms
```

