---
title: "Lab 07: Modelling course evaluations (Pt. 2)"
author: ""
date: "2021-11-12"
output: 
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.35, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Agenda

1. Lab 07: Modelling course evaluations
2. Getting started with lab


---

## Reminder: Data come from...

"Beauty in the classroom: instructors’ pulchritude and putative pedagogical productivity" 

Daniel S. Hamermesh, Amy Parker, Beauty in the classroom: instructors pulchritude and putative pedagogical productivity, Economics of Education Review, Volume 24, Issue 4, August 2005. http://www.sciencedirect.com/science/article/pii/S0272775704001165

---

## Some notes on this lab
- This is an extension of lab04. It may be worth briefly looking over that lab and/or the answer key to get that fresh in your mind prior to beginning this lab.
- There are three parts. We'll be grading to see that *at least one model with more than one predictor* has been *fit and interpreted*.




