---
title: "Lab 08: Logistic Regression"
author: ""
date: "2021-11-19"
output: 
  xaringan::moon_reader:
    css: "lab-slides.css"
    logo: img/cogs137-logo-hex.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.35, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
```

## Agenda

1. Lab 08: Modelling resumes
2. Getting started with lab


---

## Data come from...

“Are Emily and Greg More Employable than Lakisha and Jamal? A Field Experiment on Labor Market Discrimination.”

Bertrand, Marianne, and Sendhil Mullainathan. 2003. [https://doi.org/10.3386/w9873](https://doi.org/10.3386/w9873).

---

## Study

Goal: understand the influence of race and gender on job application callback rates

--

#### Specs:
- Two cities: Boston and Chicago
- Time: several months in 2001 and 2002

--

#### Plan: 

1. Researchers generated resumes, randomizing years of experience and education details
2. Then: randomly assigned a name to the resume that would communicate the applicant's gender and race 
  - they tested these names and removed those that did not suggest gender and race consistently
  - i.e. Lakisha was a name that their survey indicated would be interpreted as a black woman, while Greg was a name that would generally be interpreted to be associated with a white male."

---

## Backwards elimination (Logistic Regression)

- Start with **full** model (including all candidate explanatory variables and all
candidate interactions)

- Remove one variable at a time, and select the model with the *lowest* AIC

- Continue until AIC does not decrease

--

- You do NOT have to include every model in your lab...just the final one you settle on.


---

## A note on this lab
- There are three parts. We'll be grading to see that you've done *some* EDA and have fit and interpreted at least two models (single and multiple predictors model)
- It's OK if you don't get to backwards elimination


